### 2014-09-20 paper.tex
updated discussion, bitscores part and others.

% @CTB is Oases-M meant to work with non-oases-assembled transcripts? is it a
% general merging tool?
% @LP I think it's kind of a general merging tool.

@@

Of 105,461 splice junctions from the gene models, 2,057 (2\%) junctions have no
spliced reads and only 4,626 (4.4\%) junctions have fewer than 4 spliced reads
(Fig.~\ref{cdf_splice}).  Note that 710 junctions are in chrUn\_random contigs,
which may have a great number of genome missassemblies. The number of junctions
outside random chromosomes with no reads is 1,347/99,986 (1.3\%).
% @CTB how do you get junctions with no spliced reads!?
% @LP our models have many 2,400 more junctions in random chromosomes than
% cufflinks and 710 of no reads junctions are from random chromosomes. It is
% likely that those junctions are from missassemblies.
% @LP Another reason is it has something to do with variation in read
% mapping between Tophat and bowtie.  Tophat maps spliced reads with less
% stringency because it has to allow gap alignment and it uses -v mode to map
% spliced reads. On the other hand, the default mode for bowtie is -n mode,
% which is the one I used.
% @LP I used a parameter that set max alignments to 100, only one alignment will
% be reported if there are more than 100 alignments. Maybe I should let
% bowtie report all alignments instead.
More than 95\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?!
% @LP the reason I use 4 as a cutoff because there are four samples and if at
% least one read from each sample map to the junction, it could be real.
% Apparently, it is not always the case, but that is my simple logic.

@@

As shown above, Cufflinks and the Gimme method detect a number of distinct but
equally valid splice junctions, which suggests that we could obtain greater
sensitivity to exon-exon junctions in our gene models by merging both sets of
predictions.  We therefore used Gimme to incorporate the Cufflinks models to
global and local assembly, Table.~\ref{genes_transcripts}.  This resulted in a
decreased number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{}). The merged gene models recover 49.3\% and 56.2\% of
splice junctions from ESTs and ENSEMBL respectively
(Figure~\ref{combined_venn}), which is about 10\% greater than that from
corresponding unmerged gene models.

@@

To validate our predicted isoforms, we extracted putative coding sequences from
our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan successfully
translated 28,772 of 34,800 (82.6\%) of our isoforms to protein sequences with
50 or more amino acids.  We then searched for homologous sequences in mouse
ENSEMBL, and found that 22,991 (79.9\%) of our isoforms from 12,945 distinct
genes match mouse proteins at a bit score $\ge1.0$.
These matches have a bit-score/length ratio greater than 1, which indicates a
good agreement between chicken and mouse proteins.

% @CTB for discussion: what's surprising about all of this? so many
% splice variants, etc?  Note we need to go over pickrell paper carefully.

@@

For organisms with a reference genome and Ensembl annotation, building gene
models from RNA-Seq reads using Cufflinks with Ensembl gene models as a
reference guide seems to be a preferable method for gene and isoform expression
analysis.  However, we find that Ensembl models in chicken are missing a
substantial number of splice variants found in ESTs and mRNAs, and Cufflinks does
not recover a number of them.  A previous study by Schulz \emph{et
al.}\cite{Schulz:2012je} demonstrated that Cufflinks outperformed \emph{de novo}
assembly in detecting genes and isoforms in Ensembl annotations in mouse.
We find that Cufflinks almost exclusively finds splice junctions in
mouse Ensembl gene models, whereas our gene models include splice junctions that are not in
Ensembl gene models but that {\em are} well supported by ESTs (Figure~\ref{mus_venn}).

To build mouse Cufflinks gene models, we used Tophat default parameters claimed
to be fine-tuned for mammalian RNA-Seq reads, but Cufflinks still poorly
detected splice junctions not in Ensembl.  This suggests that the parameters may
have been set based on mammalian Ensembl models, which could limit the
efficiency of finding novel splice variants in other non-mammal organisms.  In
contrast, {\em de novo} assembly does not rely on species-specific parameters.
We therefore recommend that a combination of Cufflinks models and
transcripts from {\em de novo assembly} be used to build gene models that
include more splice variants.

We have also shown that the local assembly could greatly increase sensitivity
of splice variant detection. However, how the local assembly affects the
assembly process is not clearly understood.  We speculate that reads with
multiple alignments may play a significant role in enhancing the assembly of
splice variants.  Understanding of the mechanisms may lead to a technique that
can be applied to find splice variants in organisms that lack a reference
genome.

Both Cufflinks and our pipeline rely on a reference genome; therefore,
the quality of the genome will greatly affect the quality of the gene models.
In this study gene models were built from chicken genome version 2.1 (galGal3),
which contains ~17 Mb of sequence duplications and missassemblies
that were eliminated in the latest version of genome assembly
(galGal4).  Duplications and misassemblies lead to false splice junctions, which
in turn produce a large number of splice variants as observed in some
chromosomes.

@@

\subsection*{Quality trimming of reads}
Both single- and paired-end reads in this study were trimmed using Condetri
version 2.1 with default parameters.  In addition, the first 10 bases of each
reads were trimmed off due to an inconsistency of base-calling.

% (as shown in supplementary figure S?.)

@@

The mouse RNA-Seq dataset (SRX062280) was downloaded from the Short Read
Archive.  Chicken RNA-Seq datasets were obtained from
sequencing of mRNAs from spleen of chicken line 6 and 7.

@@

Single and paired-end reads were mapped to the chicken genome by
Tophat2\cite{Trapnell:2009dp} release 2.0.0 using default parameters without
annotations.  All reads were mapped to cDNA sequences derived from gene models
by Bowtie2\cite{Langmead:2009fv} with default parameters \texttt{(n=2, l=28,
e=70, k=1)}.  Reads from the mouse dataset were mapped to the mouse genome
(mm9) downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.

@@

Reads from each dataset were first assembled separately by global assembly
without using a reference genome.  In contrast, reads from each dataset were
first mapped to the chicken genome using Tophat2.  Then only reads mapped to
the genome were assembled by chromosome in the local assembly
(Fig.~\ref{local_assembly}).  Global and local assembly was performed using
Velvet version 1.2.03\cite{Zerbino:2008vu} with default parameters except for
hash length (k-mer).  A range of k-mer length from 21-31 was used to assemble
reads from chicken data and k-mer length 27 was used to assemble reads from
mouse data.  Lastly, transcripts from both methods were assembled by Oases
version 0.2.06\cite{Schulz:2012je}.  A poly-A tail, short transcripts and
transcripts with low complexity are removed by seqclean~\cite{seqclean} with
default parameters.  Redundant transcripts are removed by cd-hit-est from the
CD-HIT suite\cite{Li:2006hr}.  A substantial number of transcripts are removed
at this step, which facilitates gene model construction process.

We obtained 339,199 transcripts, of which 315,998 transcripts (93.2\%) mapped
to chicken genome.  Only transcripts mapped to chicken genome are used to build
gene models.

@@

Figure~\ref{overall_pipeline} depicts an overall gene model construction
pipeline.  Transcripts of all datasets from local and global assembly were
mapped to the chicken genome using BLAT\cite{Kent:2002tv} \texttt{(-t=dna -q=dna
-noHead -out=psl -mask=lower -extendThroughN -dots=1000)}.  Alignments and gaps
from BLAT outputs are considered exons and introns respectively.  Optionally,
data from other sources (ESTs, RefGenes, etc.) can be incorporated with
transcripts from the assembly to improve gene models.  All transcripts are then
assembled using Gimme, a program that assembles transcripts based on their
alignments to the reference genome.  An algorithm for assembling transcripts is
described below.  A maximum set of transcripts obtained from Gimme are then
reduced to only a minimum set of transcripts that contain all splice junctions
and untranslated regions (UTRs).  After that, transcripts that are highly
similar ($>99\%$) are clustered and removed by CD-HIT version
4.5.6\cite{Li:2006hr}.  Parameters for clustering are \texttt{-n 10, -r 0 and -c
0.99}.  Only a representative of each cluster is kept in gene models.

@@

The algorithm first builds an intron graph using introns as nodes.  Each intron
contains exons one of whose splice sites perfectly match intron boundaries.
Exons are considered incomplete and eliminated if they locate at the $3'$ or
$5'$ end of the transcripts and they are not the largest exons.
% @CTB ???
  ,
Fig.~\ref{algorithm}).  Transcripts were then grouped into the same gene if
they have at least one intron or exon in common.  Then, a splice graph composed
of exons is created and structures of isoforms are derived from traversing
paths in the splice graph.  Gimme is open-source and available at
\texttt{https://github.com/ged-lab/gimme}.

@@

To identify unique sequences from two datasets, a set of 20-mers is created for
both datasets using khmer\cite{khmer} (@CTB what script? I may have just removed it :).  Then, 20-mers from a query dataset are
compared with 20-mers from the target dataset.  The sequence is considered
unique if more than 90\% of 20-mers in the query is unique.  Any unique region
shorter than 300 bp is ignored.

@@

>>>>

% @CTB is Oases-M meant to work with non-oases-assembled transcripts? is it a
% general merging tool?
% @LP I think it's kind of a general merging tool based on de Bruijn graph.

@@

Of 105,461 splice junctions from the gene models, 2,057 (2\%) junctions have no
spliced reads and only 4,626 (4.4\%) junctions have fewer than 4 spliced reads
(Fig.~\ref{cdf_splice}).  Note that 710 junctions are in chrUn\_random contigs,
which may have a great number of genome missassemblies. The number of junctions
outside random chromosomes with no reads is 1,347/99,986 (1.3\%).
% @CTB how do you get junctions with no spliced reads!?
% @LP our models have many 2,400 more junctions in random chromosomes than
% cufflinks and 710 of no reads junctions are from random chromosomes. It is
% likely that those junctions are from missassemblies.
% @LP Another reason is it has something to do with variation in read
% mapping between Tophat and bowtie.  Tophat maps spliced reads with less
% stringency because it has to allow gap alignment and it uses -v mode to map
% spliced reads. On the other hand, the default mode for bowtie is -n mode,
% which is the one I used.
% @LP I used a parameter that set max alignments to 100, only one alignment will
% be reported if there are more than 100 alignments. Maybe I should let
% bowtie report all alignments instead.
More than 95\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?!
% @LP the main reason I use four reads as a cutoff because there are four
% samples and if at least one read from each sample map to the junction, it
% will likely be real.  Apparently, it is not always the case, but that is just
% my simple logic.

@@

As shown above, Cufflinks and the Gimme method detect a number of distinct but
equally valid splice junctions, which suggests that we could obtain greater
sensitivity to exon-exon junctions in our gene models by merging both sets of
predictions.  We therefore used Gimme to incorporate the Cufflinks models to
global and local assembly, Table.~\ref{genes_transcripts}.  This resulted in a
decreased number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{}). The merged gene models recover 49.3\% and 56.2\% of
splice junctions from ESTs and ENSEMBL respectively
(Fig.~\ref{combined_venn}), which is about 10\% greater than that from
corresponding unmerged gene models.

@@

% @LP Maybe I should add bitscores from Cufflinks so that we can compare with
% those from Gimme. If the results are similar, we can say that extending
% Cufflinks with assembly does not affect gene structures in a bad way.
To validate our predicted isoforms, we extracted putative coding sequences from
our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan successfully
translated 28,772 of 34,800 (82.6\%) of our isoforms to protein sequences with
50 or more amino acids.  We then searched for homologous sequences in mouse
ENSEMBL, and found that 22,991 (79.9\%) of our isoforms from 12,945 distinct
genes match mouse proteins at a bit-score/length $\ge1.0$
(Fig.~\ref{bitscore}).  A bit-score/length ratio greater than 1 indicates that
approximately more than half of the protein sequence matches mouse proteins
well. The results suggest that merging gene models from Cufflinks and {\em de
novo} aseembly did not disrupt the structure of protein sequences.

% @CTB for discussion: what's surprising about all of this? so many
% splice variants, etc?  Note we need to go over pickrell paper carefully.
% @LP I discuss about it below. Hope it helps.

@@

\subsection*{{\em De novo} assembly should be used to extend gene models from
Cufflinks}
For organisms with a reference genome and ENSEMBL annotation, Cufflinks with
ENSEMBL gene models as a reference guide seems to be a preferable method for
building gene models. However, we have shown that ENSEMBL models in chicken are
missing a substantial number of splice variants found in ESTs and mRNAs that
are also not recovered by Cufflinks. Moreover, results from mouse data suggest
that Tophat's default parameters may have been tailored to maximize the finding
of isoforms in ENSEMBL models (Fig.~\ref{mus_venn}), which could lead to
restricted detection of novel splice variants.
% 85,512 (97.2\%) of splice
% junctions found by Cufflinks are also canonical compared to 92.6\% in mouse
% ENSEMBL gene models. Assuming that mouse data were not selected for splice
% variants with canonical splice junctions, the percentage of canonical splice
% sites found by Cufflinks is rather high. 
Therefore, gene models built from RNA-Seq reads using Cufflinks + ENSEMBL gene
models alone may not be suitable for genes and isoforms analyses that include
finding novel genes and splice variants.  {\em De novo} assembly is not tied to
any specific gene model parameters and does not only detect canonical splice
sites, thus we recommend that it is used to extend gene models built from
Cufflinks + ENSEMBL gene models, especially in organisms with relatively
incomplete ENSEMBL gene models.  Besides {\em de novo} assembly,
annotation from other sources can also be incorporated to extend Cufflinks models
using our pipeline.

% Moreover, efficiency of finding minor and non-canonical splice sites
% of Cufflinks is also low, which could cause a problem when Cufflinks is used to
% construct gene models for organisms with more isoforms with minor and
% non-canonicla splice sites.

\subsection*{Reads with multiple alignments may be key to detection of unique
splice variants from local assembly}
We have shown that local assembly greatly increases sensitivity of splice
variant detection; however, the mechanism is still not clearly understood.  In
local assembly, reads mapped to each chromosome are assembled separately and
because Tophat reports multiple alignments for spliced reads, a single spliced
read from an exon junction present in multiple chromosomes gets assembled
multiple times. We speculate that this increases the coverage of exon junctions
found in multiple chromosomes, which help improve the assembly of splice
variants with low coverage. Further investigation of an inherent mechanism of
local assembly may lead to a method that increases sensitivity of splice
variants in organisms that lack a reference genome.

\subsection*{Spurious predicted isoforms are due to poor genome assembly}
Both Cufflinks and our pipeline rely on a reference genome; therefore, the
quality of the genome greatly affects the quality of the gene models.  In this
study, gene models were built from chicken genome version 2.1 (galGal3), which
contains ~17 Mb of sequence duplications and missassemblies that were
eliminated in the latest version of genome assembly (galGal4).  Duplications
and misassemblies lead to spurious splice junctions, which in turn produce
spurious splice variants. Figure~\ref{bitscore} shows that the median of
bitscore/length ratios of isoforms is lower than that of genes, suggesting
poorer alignments of isoforms to mouse protein sequences.

@@

\subsection*{Quality trimming of reads}
Both single- and paired-end reads in this study were trimmed using Condetri
version 2.1 with default parameters.  In addition, the first 10 bases of each
reads were trimmed off due to an inconsistency of base-calling as shown in
supplementary Figure 1.

@@

The mouse RNA-Seq dataset (SRX062280) was downloaded from the Short Read
Archive.  Chicken RNA-Seq datasets were obtained from sequencing of mRNAs from
spleens of control and infected chicken (4 days post infection) lines 6 and 7
using standard Illumina protocol for non-strand specific single-end reads.

@@

Single and paired-end reads were mapped to the chicken genome by
Tophat\cite{Trapnell:2009dp} release 1.3.1 using default parameters without
annotations.  All reads were mapped to cDNA sequences derived from gene models
by Bowtie\cite{Langmead:2009fv} release 1.0.0 with default parameters Reads from the
mouse dataset were mapped to the mouse genome (mm9) downloaded from Tophat
website \texttt{http://tophat.cbcb.umd.edu}.

@@

% @LP check the version of tophat
Reads from each dataset were first assembled separately by global assembly
without using a reference genome.  In contrast, reads from each dataset were
first mapped to the chicken genome using Tophat2.  Then only reads mapped to
the genome were assembled by chromosome in the local assembly
(Fig.~\ref{local_assembly}).  Global and local assembly was performed using
Velvet version 1.2.03\cite{Zerbino:2008vu} with default parameters except for
hash length (k-mer).  A range of k-mer length from 21-31 was used to assemble
reads from chicken data and k-mer length 27 was used to assemble reads from
mouse data.  Lastly, transcripts from both methods were assembled by Oases
version 0.2.06\cite{Schulz:2012je}.

A poly-A tail, short transcripts and transcripts with low complexity are
removed by SeqClean~\cite{seqclean} with default parameters.  Redundant
transcripts are removed by \texttt{cd-hit-est} from the CD-HIT
suite\cite{Li:2006hr} release 4.5.4.  A substantial number of transcripts are
removed at this step, which facilitates gene model construction process.  We
obtained 339,199 transcripts, of which 315,998 transcripts (93.2\%) mapped to
chicken genome.  Only transcripts mapped to chicken genome are used to build
gene models.

@@

Figure~\ref{overall_pipeline} depicts an overall gene model construction
pipeline.  Transcripts of all datasets from local and global assembly were
mapped to the chicken genome using BLAT\cite{Kent:2002tv}. Alignments and gaps
from BLAT outputs are considered exons and introns respectively.  Optionally,
data from other sources (ESTs, RefGenes, etc.) can be incorporated with
transcripts from the assembly to improve gene models.  All transcripts are then
assembled using Gimme, a program that assembles transcripts based on their
alignments to the reference genome.  An algorithm for assembling transcripts is
described below.  A maximum set of transcripts obtained from Gimme are then
reduced to only a minimum set of transcripts that contain all splice junctions
and untranslated regions (UTRs).

@@

The algorithm first builds an intron graph using introns as nodes.  Each intron
contains exons whose splice sites perfectly match intron boundaries. At the
$5\prime$ and the $3\prime$ ends, some exons may be smaller than others due to
incomplete assembly (supp. Fig.??). Only the largest exons are kept to ensure
that the transcript has the most complete exons (Fig.~\ref{algorithm}).
Transcripts were then grouped into the same gene if they have at least one
intron or exon in common.  Then, a splice graph composed of exons is created
and structures of isoforms are derived from traversing paths in the splice
graph.  Gimme is open-source and available at
\texttt{https://github.com/ged-lab/gimme}.

@@

To identify unique sequences from two datasets, a set of 20-mers is created for
both datasets using khmer\cite{} . Then, 20-mers from a query dataset are
compared with 20-mers from the target dataset.  The sequence is considered
unique if more than 90\% of 20-mers in the query is unique.  Any unique region
shorter than 300 bp is ignored.
% @CTB what script? I may have just removed it.
% @LP The script is named assembly-diff.py. I have it in my repo. It is a
% modified version of yours.

% @LP how do I cite khmer? Should I cite QP paper?

@@

\subsection*{Pipeline for the study}
The pipeline, scripts and instructions for reproducing this study are open
source and available at \texttt{http://github.com/likit/gimme-paper}.
